---
title: "Agglomeration"
format:
  revealjs:
    self-contained: true
    slide-number: true
    preview-links: auto
    logo: images/mia_logo.png
    footer: <https://microbiome.github.io/>
date: last-modified
date-format: full
bibliography: references.bib
---

## Agglomeration

_Combining features into higher-level taxa_

## Why?

- Focus on biologically meaningful units
- Reduce sparsity and noise
- Improve interpretability of downstream analyses

::: {.notes}

- Focusing on higher-level taxa (e.g., genera instead of thousands of ASVs) helps highlight biologically relevant patterns.
- 16S sequencing often does not resolve taxa below genus level reliably.
- Agglomeration reduces noise from sequencing artifacts and rare features.
- Makes downstream analyses like diversity metrics, ordination, and association studies more interpretable.
- Note: Some resolution is lost when collapsing features—balance simplicity with detail.

:::

##

```{r}
#| label: agglomeration

library(ggplot2)
library(ggalluvial)

df <- data.frame(
  Species = c("Species 1","Species 2","Species 3","Species 4","Species 5"),
  Genus   = c("Genus A","Genus A","Genus A","Genus B","Genus B"),
  Count   = c(3,5,4,4,6)  # example abundances
)

ggplot(df,
       aes(axis1 = Species, axis2 = Genus, y = Count)) +
  geom_alluvium(aes(fill = Genus), alpha = 0.6, color = "grey40") +
  geom_stratum(width = 0.3, fill = "grey90", color = "black") +
  # Add counts to stratum labels
  geom_text(stat = "stratum",
            aes(label = paste0(after_stat(stratum), "\n(", after_stat(count), ")")),
            size = 5) +
  scale_x_discrete(limits = c("Species","Genus"), expand = c(.1, .1)) +
  theme_void(base_size = 16) +
  theme(legend.position = "none")

```

::: {.notes}

Species belonging to Genus A are summed up together...

:::

## Demonstration

```r
library(mia)

tse <- agglomerateByRank(tse, rank = "Phylum")
print(tse)
```

```{r}
#| label: agglomerate_rank

library(mia)
data(GlobalPatterns)
tse <- GlobalPatterns
tse <- agglomerateByRank(tse, rank = "Phylum")
tse
```

## {auto-animate="true"}

```r
tse <- agglomerateByRanks(tse)
print(tse)
```
```{r}
#| label: agglomerate_all_ranks

library(mia)
data(GlobalPatterns)
tse <- GlobalPatterns
tse <- agglomerateByRanks(tse)
tse
```

## Exercises

From [OMA online book, Chapter 11: Agglomeration](https://microbiome.github.io/OMA/docs/devel/pages/agglomeration.html)

- 1.2, 1.3, 1.4, 1.5, 1.6

## References
