---
title: "Agglomeration"
subtitle: "Combining features into higher-level taxa"
format:
  revealjs:
    theme: [default, bioconductor.scss]
    transition: fade
    transition-speed: fast
    self-contained: true
    slide-number: true
    preview-links: auto
    logo: images/mia_logo.png
    footer: <https://microbiome.github.io/>
date: last-modified
date-format: full
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: ragg_png
---

## Why agglomerate?

-   Focus on biologically meaningful units
-   Reduce sparsity and noise
-   Improve interpretability of downstream analyses

::: notes
-   Focusing on higher-level taxa (e.g., genera instead of thousands of ASVs) helps highlight biologically relevant patterns.
-   16S sequencing often does not resolve taxa below genus level reliably.
-   Agglomeration reduces noise from sequencing artifacts and rare features.
-   Makes downstream analyses like diversity metrics, ordination, and association studies more interpretable.
-   Note: Some resolution is lost when collapsing features—balance simplicity with detail.
:::

## Summing up the taxonomy

```{r}
#| label: agglomeration
#| fig-width: 8
#| fig-height: 4.5

library(ggplot2)
library(ggalluvial)

df <- data.frame(
    Species = c("Species 1", "Species 2", "Species 3", "Species 4", "Species 5"),
    Genus   = c("Genus A", "Genus A", "Genus A", "Genus B", "Genus B"),
    Count   = c(3, 5, 4, 4, 6)
)

ggplot(df, aes(axis1 = Species, axis2 = Genus, y = Count)) +
    geom_alluvium(aes(fill = Genus), alpha = 0.6, color = "grey40") +
    geom_stratum(width = 1/3, fill = "grey95", color = "grey30") +
    geom_text(
        stat = "stratum",
        aes(label = paste0(after_stat(stratum), "\n(", after_stat(count), ")")),
        size = 5) +
    scale_x_discrete(limits = c("Species", "Genus"), expand = c(0.1, 0.1)) +
    scale_fill_manual(values = c("#1a81a9", "#87b13f")) +
    theme_void(base_size = 16) +
    theme(legend.position = "none")
```

::: notes
Counts of species belonging to the same genus are summed up: Genus A gets 3 + 5 + 4 = 12.
:::

## Demonstration {.center}

## Agglomerate to one rank {auto-animate="true"}

```{r}
#| label: agglomerate_rank
#| echo: true
library(mia)
data("GlobalPatterns", package = "mia")
tse <- GlobalPatterns

tse_phylum <- agglomerateByRank(tse, rank = "Phylum")
c(features = nrow(tse), phyla = nrow(tse_phylum))
```

::: notes
19,216 features collapse to 66 phyla. Assays are summed and rowData is trimmed to the ranks at and above phylum.
:::

## Agglomerate to all ranks {auto-animate="true"}

```{r}
#| label: agglomerate_all_ranks
#| echo: true
tse <- agglomerateByRanks(tse)
sapply(altExps(tse), nrow)
```

::: notes
agglomerateByRanks stores every rank as an alternative experiment, e.g. altExp(tse, "Genus").
:::

## Exercises

::: callout-tip
## OMA online book, Chapter 10: Agglomeration

[microbiome.github.io/OMA › Agglomeration](https://microbiome.github.io/OMA/docs/devel/pages/agglomeration.html)

Exercises **1.2, 1.3, 1.4, 1.5, 1.6**
:::

Try it on real data: [GuptaA_2019 example data](https://microbiome.github.io/outreach/gupta2019.html)
