---
title: "Basic Exploration"
format: html
editor: visual
---

## Overview

The following packages are needed to succesfully run the examples in this notebook:

-   mia: tools for microbiome data analysis

-   scater: plotting data from TreeSummarizedExperiments

```{r}
#| label: setup
#| include: false
library(mia)
library(scater)

knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## TreeSummarizedExperiment

To get started, we import Tengeler2020 from the mia package and store it into a variable, on which we will work for the rest of the tutorial.

```{r}
#| label: import-tengeler

# load dataset and store it into tse
data("Tengeler2020", package = "mia")
tse <- Tengeler2020
```

```{r}
tse
```

```{r}
dim(tse)
```

```{r}
colnames(tse)
```

```{r}
rownames(tse)
```

## Assays

```{r}
assays(tse)
```

```{r}
head(assay(tse, "counts"))
```

## colData

```{r}
names(colData(tse))
```

## rowData

```{r}
names(colData(tse))
```

```{r}

```

## Other elements

```{r}
altExp(tse, "my_alt_exp") <- tse[1:10, ]
altExps(tse)
```

```{r}
rowTree(tse)
```

```{r}
reducedDims(tse)
```

```{r}
metadata(tse)
```
