---
title: "GuptaA_2019"
subtitle: "Example data: colorectal cancer gut metagenomes from India"
format:
  revealjs:
    theme: [default, bioconductor.scss]
    transition: fade
    transition-speed: fast
    self-contained: true
    slide-number: true
    preview-links: auto
    logo: images/mia_logo.png
    footer: <https://microbiome.github.io/>
date: last-modified
date-format: full
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: ragg_png
---

## The study

Gut microbiome of colorectal cancer (CRC) patients in India [@Gupta2019]

-   **30 CRC patients** and **30 healthy controls** from Bhopal and Kerala
-   Controls from an earlier study of healthy Indians [@Dhakan2019]
-   Stool shotgun metagenomes, profiles from
    [curatedMetagenomicData](https://bioconductor.org/packages/curatedMetagenomicData)

::: notes
Gupta et al. associated the flavonoid-degrading bacterium Flavonifractor plautii with colorectal cancer in Indian patients. This data set is useful for teaching because it is small (60 samples), has a clear case–control design, and includes a phylogenetic tree.
:::

## Import the data

Files and import script: [github.com/microbiome/data › GuptaA_2019](https://github.com/microbiome/data/tree/main/GuptaA_2019)

```{r}
#| label: import
#| echo: true
library(mia)
library(ape)

url <- "https://raw.githubusercontent.com/microbiome/data/main/GuptaA_2019"
read <- function(f) read.csv(file.path(url, f), row.names = 1)

tse <- TreeSummarizedExperiment(
    assays  = list(counts = as.matrix(read("taxonomy_abundance.csv"))),
    rowData = read("taxonomy_table.csv"),
    colData = read("sample_metadata.csv"),
    rowTree = read.tree(file.path(url, "phylogeny.tree"))
)
```

::: notes
The same files can be downloaded and read from a local folder. The repository's import.R also reads pathway abundances into a MultiAssayExperiment.
:::

## The data object {.smaller}

```{r}
#| label: object
#| echo: true
tse
```

## Samples

::: columns
::: {.column width="50%"}
```{r}
#| label: tab-cohort
tab <- table(tse$study_condition, tse$location)
rownames(tab) <- c("Control", "CRC")
knitr::kable(tab)
```
:::

::: {.column width="50%"}
```{r}
#| label: tab-sex
tab <- table(tse$study_condition, tse$gender)
rownames(tab) <- c("Control", "CRC")
colnames(tab) <- c("Female", "Male")
knitr::kable(tab)
```
:::
:::

Age `r min(tse$age)`–`r max(tse$age)` years (median `r median(tse$age)`);
CRC stage, location and TNM available for patients.

::: callout-warning
Location and study are partly confounded with disease status: most controls come from Bhopal, and controls and patients come from different studies.
:::

## Composition at a glance

```{r}
#| label: phyla
#| fig-width: 9
#| fig-height: 4.2
library(ggplot2)

tse_phylum <- agglomerateByRank(tse, rank = "phylum")
tse_phylum <- transformAssay(tse_phylum, method = "relabundance")

df <- meltSE(tse_phylum, assay.type = "relabundance",
             add.col = "study_condition")
top <- names(sort(rowMeans(assay(tse_phylum, "relabundance")),
                  decreasing = TRUE))[1:5]
df$phylum <- ifelse(df$FeatureID %in% top, as.character(df$FeatureID), "Other")
df$phylum <- factor(df$phylum, levels = c(top, "Other"))
df$study_condition <- factor(df$study_condition,
                             levels = c("control", "CRC"),
                             labels = c("Control", "CRC"))

ggplot(df, aes(x = SampleID, y = relabundance, fill = phylum)) +
    geom_col(width = 1) +
    facet_grid(~ study_condition, scales = "free_x", space = "free_x") +
    scale_fill_manual(values = c("#1a81a9", "#87b13f", "#f2b134",
                                 "#e76f51", "#6a4c93", "grey75")) +
    scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
    labs(x = "Samples", y = "Relative abundance", fill = "Phylum") +
    theme_minimal(base_size = 15) +
    theme(axis.text.x = element_blank(), panel.grid = element_blank())
```

## Try it

Practise the workflow on this data set:

-   [Agglomeration](https://microbiome.github.io/outreach/agglomeration.html)
-   [Transformation](https://microbiome.github.io/outreach/transformation.html)
-   [Alpha diversity](https://microbiome.github.io/outreach/alpha_diversity.html)
    and [beta diversity](https://microbiome.github.io/outreach/beta_diversity.html):
    do CRC patients and controls differ?

## References

::: {#refs}
:::
