---
title: "Differential Abundance"
format: html
editor: visual
---

```{r}
#| label: setup
#| include: false
library(mia)
library(MicrobiomeStat)
library(tidyverse)

knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Overview

Differential Abundance Analysis (DAA) is used to identify taxa that are significantly more or less abundant in the condition compared to control. For more details, read [this chapter](https://microbiome.github.io/OMA/differential-abundance.html#differential-abundance) of the OMA book.

Many methods are available including:

-   ALDEx2

-   ANCOMBC

-   LinDA

A few things to keep in minds when performing DAA involve:

-   DAA software normally takes the counts assay as input, because they apply normalisation suitable for count data

-   DAA results will be more reproducible if the extremely rare taxa and singletons are removed in advance

-   It is recommended to run different methods on the same data and compare the results

## Preparing for DAA

Before performing DAA, it is important to agglomerate to a meaningful taxonomic rank and select only taxa above a certain prevalence and detection threshold, as this has been shown to make results more reproducible.

```{r}
data("Tengeler2020", package = "mia")
tse <- Tengeler2020
```

```{r}
tse_genus <- agglomerateByPrevalence(tse,
                                     rank = "Genus",
                                     detection = 0.001,
                                     prevalence = 0.1)
```

## Performing DAA

For this tutorial, we run the LinDA method. We first extract the counts assay and convert it into a data frame.

```{r}
otu.tab <- assay(tse_genus, "counts") |>
  as.data.frame()
```

We also need to select the columns of the colData which contain the independent variables you want to include in the model.

```{r}
meta <- colData(tse) |>
  as.data.frame() |>
  select(patient_status, cohort)
```

We are ready to run LinDA, which takes the assay count (`otu.tab`) and the variable arrays (`meta`). A formula for the model with main independent variable + covariates should be defined. The other arguments are optional but good to know.

```{r}
res <- linda(otu.tab,
             meta,
             formula = "~ patient_status + cohort", 
             alpha = 0.05, 
             prev.filter = 0, 
             mean.abund.filter = 0,
             feature.dat.type = "count")
```

## Interpreting results

Finally, we select significantly DA taxa and list it in @tbl-daa.

```{r}
#| label: tbl-daa
#| tbl-cap: DA bacterial genera. If stat > 0, abundance is higher in control, otherwise it is higher in ADHD.
signif_res <- res$output$patient_statusControl |>
  filter(reject) |>
  select(stat, padj) |>
  arrange(padj)

knitr::kable(signif_res)
```

Good job reading this tutorial. Now go [this chapter](https://microbiome.github.io/OMA/differential-abundance.html#differential-abundance) of the OMA book and try out other DAA methods on Tengeler2020.
