---
title: "A Brief Introduction to Tengeler2020"
format: revealjs
editor: visual
smaller: true
---

```{r}
#| label: setup
#| include: false
library(mia)
data("Tengeler2020", package = "mia")
tse <- Tengeler2020
```

## Study design {.scrollable}

. . .

Tengeler2020 is a TreeSummarizedExperiment from a study on the effects of gut microbiome on Attention-deficit/hyperactivity disorder (ADHD) in humanised mice.

. . .

::: columns
::: {.column width="48%"}
Young, male, germ-free C57BL/6JOlaHsd mice (n = 27) were humanised with gut microbiome from either ADHD patients (n = 3) or healthy controls (n = 3) of matched age. Notably, the mice belonged to three different cohorts.
:::

::: {.column width="52%"}
```{r}
#| label: tab-pop
tab <- table(colData(tse)$patient_status, colData(tse)$cohort)
colnames(tab) <- c("Cohort1", "Cohort2", "Cohort3")
knitr::kable(tab)
```
:::
:::

Microbiome data was obtained by 16S rRNA gene sequencing of bacterial DNA sampled from faecal pellet on a weekly basis. Then, sequence reads were assembled into Operational Taxonomic Units (OTUs) with an NG-Tax pipeline.

## Tutorials

-   [Alpha Diversity](https://microbiome.github.io/outreach/tengeler_alpha_diversity.html)

-   [Beta Diversity with Principal Coordinate Analysis](https://microbiome.github.io/outreach/tengeler_pcoa.html)

-   [Beta Diversity with Redundance Analysis](https://microbiome.github.io/outreach/tengeler_dbrda.html)

-   [Community Composition](https://microbiome.github.io/outreach/tengeler_community_composition.html)

# The end
