---
title: "Reproducible Reporting with Quarto"
format: revealjs
editor: visual
smaller: true
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE)
```

## Example 1.1: \# Markdown Syntax

In Markdown syntax, you can do the following and more.

-   Headings: `# My title`, `## My subtitle`, `### My subsubtitle`
-   Unordered lists:
    -   `-   item 1`
    -   `-   item 2`
-   Ordered lists:
    -   `1.  item 1`
    -   `2.  item 2`
-   Font: \**italic*\*, \*\***bold**\*\*, \``monospaced`\`
-   Links: `[text](url)`
-   Cross-references: `@chunk-label`

## Example 1.2: Running Code

Add code chunks with `alt + cmd + i` and click **Render** to generate a report with both text and code output.

```{r}
# R code
citation("mia")
```

## Exercise 1

-   new document: [exercise 2.1](https://microbiome.github.io/OMA/docs/devel/pages/98_exercises.html#new-document){target="_blank"}
-   code chunks: [exercise 2.2](https://microbiome.github.io/OMA/docs/devel/pages/98_exercises.html#code-chunks){target="_blank"}

## Example 2.1: knitr options

You can add options to code chunks to change their behaviour.

-   Original chunk

```{r}
print("I love my microbiome")
```

-   After adding `#| echo: false`

```{r}
#| echo: false
print("I love my microbiome")
```

-   After adding `#| eval: false`

```{r}
#| eval: false
print("I love my microbiome")
```

-   After adding `#| code-fold: true`

```{r}
#| code-fold: true
#| code-summary: "Show code"
print("I love my microbiome")
```

-   After adding `#| include: false`

```{r}
#| include: false
print("I love my microbiome")
```

## Example 2.2: More about knitr options

If you want an option to affect all chunks in a script, you can set it globally.

```{r}
#| label: global-knitr-example
#| eval: false
# Turn off chunk visibility and warnings
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
```

You can label figures (or tables) with `#| label: fig-name` and cross-reference
them with `@fig-name` (@fig-example).

```{r}
#| label: fig-example
#| fig-cap: A boxplot of the sepal length distribution by species.
data(iris)
boxplot(Sepal.Length ~ Species, data = iris)
```

## Example 2.3: YAML Parameters

At the beginning of any Quarto document, there is a box delimited by `---`.
There you can define document metadata, such as title, author, date, output
format, bibliography, citation style, theme, font size and many others.

`---`

title: "Around the gut in 24 hours"

format: html

editor: visual

smaller: true

author: Escherichia coli

date: `r Sys.Date()`

`---`

## Exercise 2

-   knitr options: [exercise 2.3](https://microbiome.github.io/OMA/docs/devel/pages/98_exercises.html#knitr-options){target="_blank"}
-   YAML parameters: [exercise 2.4](https://microbiome.github.io/OMA/docs/devel/pages/98_exercises.html#yaml-instructions){target="_blank"}

Extra:

-   Quarto parameters: [exercise 2.5](https://microbiome.github.io/OMA/docs/devel/pages/98_exercises.html#quarto-parameters){target="_blank"}

## Resources

-   [OMA Section - Quarto](https://microbiome.github.io/OMA/resources.html#quarto){target="_blank"}
-   [Chunk Option Catalogue](https://yihui.org/knitr/options/){target="_blank"}
-   [Cross References](https://quarto.org/docs/authoring/cross-references.html){target="_blank"}
-   [YAML Parameter Catalogue](https://quarto.org/docs/reference/formats/html.html){target="_blank"}
