---
title: "Beta Diversity"
format: html
editor: visual
---

## Overview

The following packages are needed to succesfully run the examples in this notebook:

-   mia: tools for microbiome data analysis

-   scater: plotting data from TreeSummarizedExperiments

-   patchwork: combining plots

```{r}
#| label: setup
#| include: false
library(mia)
library(scater)
library(patchwork)

knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

To get started, we import Tengeler2020 from the mia package and store it into a variable.

```{r}
#| label: import-tengeler

# load dataset and store it into tse
data("Tengeler2020", package = "mia")
tse <- Tengeler2020
```

After that, we transform the counts assay to relative abundances and store the new assay back in the TreeSE.

```{r}
#| label: transform-relabundance
tse <- transformCounts(tse,
                       method = "relabundance")
```

## Principal Coordinate Analysis (PCoA)

PCoA is a method to choose the dimensions of the data that explain most of the variance in the data. Distances between samples can be expressed by several ecological indices, such as Bray-Curtis and Aitchison dissimilarities.

Here, we run multi-dimensional scaling (a type of PCoA) with Bray-Curtis dissimilarity and based on the relabundance assay.

```{r}
#| label: run-bray
tse <- runMDS(tse,
              FUN = vegan::vegdist,
              method = "bray",
              name = "Bray",
              assay.type = "relabundance")
```

We then visualise the first two dimensions.

```{r}
#| label: plot-bray
p1 <- plotReducedDim(tse, "Bray",
                     colour_by = "patient_status",
                     shape_by = "cohort")
```

```{r}
#| label: prep-bray
#| include: false
e <- attr(reducedDim(tse, "Bray"), "eig")
rel_eig <- e / sum(e[e > 0])

p1 <- p1 + labs(x = paste("Bray 1 (", round(100 * rel_eig[[1]], 1), "%", ")", sep = ""),
                y = paste("Bray 2 (", round(100 * rel_eig[[2]], 1), "%", ")", sep = ""))
```

```{r}
#| label: fig-bray
#| fig-cap: Ordination plots based on Bray-Curtis index. Samples are coloured by patient status and shaped by cohort.
#| fig-width: 7
#| fig-asp: 0.7
p1
```

```{r}
#| label: run-unifrac
tse <- runMDS(tse,
              FUN = mia::calculateUnifrac,
              name = "Unifrac",
              tree = rowTree(tse),
              ntop = nrow(tse),
              assay.type = "counts")
```

```{r}
#| label: plot-unifrac
p2 <- plotReducedDim(tse, "Unifrac",
                     colour_by = "patient_status",
                     shape_by = "cohort")
```

```{r}
#| label: prep-unifrac
#| include: false
e <- attr(reducedDim(tse, "Unifrac"), "eig")
rel_eig <- e / sum(e[e > 0])

p2 <- p2 + labs(x = paste("Unifrac 1 (", round(100 * rel_eig[[1]], 1), "%", ")", sep = ""),
                y = paste("Unifrac 2 (", round(100 * rel_eig[[2]], 1), "%", ")", sep = ""))
```

```{r}
#| label: fig-unifrac
#| fig-cap: Ordination plot based on Unifrac index. Samples are coloured by patient status and shaped by cohort.
#| fig-width: 7
#| fig-asp: 0.7
p2
```

Finally, we combine the two plots with the patchwork syntax and compare them in @fig-beta. (Add another 2 plots?)

```{r}
#| label: fig-beta
#| fig-cap: Ordination plots based on (A) Bray-Curtis or (B) Unifrac index. Samples are coloured by patient status and shaped by cohort.
#| fig-asp: 1.4
(p1 / p2) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A")
```

It is also possible to specify the number of output dimensions with the argument `ncomponents`. Here, we show it with the UMAP ordination method.

```{r}
#| label: run-umap
tse <- runUMAP(tse,
               assay.type = "counts",
               ncomponents = 3)
```

We then plot all three dimensions, but you could also plot a pair of dimensions (1 and 3, 2 and 3 or 1 and 2) with the argument `ncomponents`.

```{r}
#| label: plot-umap
p <- plotReducedDim(tse, "UMAP",
                    ncomponents = 1:3,
                    colour_by = "patient_status",
                    shape_by = "cohort")
```

```{r}
#| label: fig-umap
#| fig-cap: UMAP plot of the first three dimensions.
#| echo: false
p
```
