---
title: "Reproducible Reporting with Quarto"
format: html
editor: visual
---

## Overview

Quarto follows the markdown and rmarkdown conventions of **literate programming**, where code and text are combined for reproducible reporting. In other words, the analytical pipeline is embedded in the report and the code is run every time the report is compiled, so that you don't have to produce results separately and then include them in a one-time-only report. This way, using Quarto assists in building a reproducible analytical worflow.

## Getting started

### Markdown sytax

The markdown syntax is a powerful that lets you add many different things to your report. Here we go through a few of the elements.

-   Titles, defined with `#`; you can also make subtitles, subsubtitles and even lower-level headings by using more hashes, so `### my subsubsection` would be 3rd-level heading

-   Lists, like this one; you can make unordered lists with `*` for each item, or ordered lists with `1.`, `2.`, `3.` and so on

-   Links, which are made with `[text](url)`

### Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

### Figures and tables

Figures and tables returned as output of your code can be easily embedded to the report and cross-referenced from other parts of the report. Here we show it with the iris dataset.

```{r}
data(iris)
```

Below is a boxplot of flower sepal length by species (see @fig-example). You can add cross-references to figures and tables with the syntax `@chunk-label`, which is further discussed in @sec-optcat. Other items such as sections, equations and publications can be referenced as well as explained [here](https://quarto.org/docs/authoring/cross-references.html).

```{r}
#| label: fig-example
#| fig-cap: This is a figure caption.
boxplot(Sepal.Length ~ Species, data = iris)
```

```{r}
#| label: tbl-example
#| tbl-cap: This is a table caption.
knitr::kable(table(iris$Species))
```

## Modifying structure and layout

### Adding knitr options

You can add options to executable code like this:

```{r}
#| label: chunk-name
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

### Option catalogue {#sec-optcat}

This are some of the knitr chunk options that I find most useful. A full list of options is available [here](https://yihui.org/knitr/options/).

-   label: gives a name to the code chunk, which comes handy for troubleshooting programs as well as adding cross-references (notice that chunks returning figures or tables need a special prefix `fig-` and `tbl-`, respectively)

-   echo / include: the former hides the code but shows the output, whereas the latter hides both the code and the output

-   eval: whether to evaluate or not the code

-   message / warning: hide messages or warnings returned by the code, respectively

-   fig / tbl: this one is not really an option, but you have to add this prefix to a label so that it is recognised as figure or table, respectively

-   fig-cap / tbl-cap: define caption for figure or table, respectively

-   fig-width + fig-asp: the former defines the width of an image, whereas the latter sets its aspect ratio so that `fig-height = fig-width * fig-asp`, so these two make a great combo to guess the correct size for your plots

-   fig.path: indicates the relative path to which figures should be saved

### Global knitr options

If you are too lazy to set knitr options for every single chunk, don't worry! knitr has the solution for you. Just specify global options for the entire document on the very top upstream like this:

```{r}
#| label: global-knitr-example

# set message and warning to false and specify folder where figures should be saved for the entire document (put upstream everything)
knitr::opts_chunk$set(message = FALSE, warning = FALSE,
                      fig.path = "my/fig/dir")
```

### YAML Parameters

At the beginning of the raw document, you'll see a box delimited by `---`. There you can define document metadata like title, author, date, but also more advanced execution options such as output format, bibliography and citation style, theme, font size and many others. A full list of available parameters can be found [here](https://quarto.org/docs/reference/formats/html.html).

## Useful Resources

We have gathered learning material about Quarto in [this chapter](https://microbiome.github.io/OMA/resources.html#quarto) of the OMA book. We encourage you to look into it.

When you feel ready for challenge, go have a look at [our exercises](https://microbiome.github.io/OMA/exercises.html#workflows) about Quarto in the OMA book. Feel free to ask if you'd like to discuss the solution strategy.
