---
title: "Microbiome data science workflow"
subtitle: "Standard components of a taxonomic profiling study"
format:
  revealjs:
    theme: [default, oulu2026.scss, workflow.scss]
    self-contained: true
    slide-number: true
    preview-links: auto
    transition: slide
    background-transition: fade
    logo: images/mia_logo.png
    footer: <https://microbiome.github.io/>
date: last-modified
date-format: full
bibliography: references.bib
execute:
  echo: true
  eval: false
---

## A typical workflow

::: flow
[Import]{.step .navy}
[Preprocess]{.step .teal}
[Alpha]{.step .teal}
[Beta]{.step .coral}
[Taxa]{.step .coral}
[Report]{.step .gold}
:::

-   Key elements of a taxonomic profiling study
-   Mixed and matched in real case studies
-   Examples for each step in [OMA](https://microbiome.github.io/OMA/)

::: notes
This session gives an overview of the key elements of a typical data
science workflow in a taxonomic profiling study. Real case studies mix
these in various ways.
:::

## Your task

::: statement
Build a clear, compact and reproducible report on your data.
:::

::: cards
::: {.card .teal}
[📓]{.icon}

[Reproducible]{.card-title}

Quarto / R Markdown report that renders from raw data
:::

::: {.card .coral}
[🎯]{.icon}

[Selective]{.card-title}

Choose summaries and analyses that answer your question
:::

::: {.card .gold}
[📚]{.icon}

[Build on OMA]{.card-title}

Reuse material and examples from the [OMA book](https://microbiome.github.io/OMA/)
:::
:::

## Setup

```{r}
#| label: setup
library(mia)
library(miaViz)
library(scater)

# Example data set: human gut microbiome with sample metadata
data("peerj13075", package = "mia")
tse <- peerj13075

# Relative abundances and CLR for downstream steps
tse <- transformAssay(tse, method = "relabundance")
tse <- transformAssay(tse, method = "clr", pseudocount = TRUE)

# Genus-level summary
tse_genus <- agglomerateByRank(tse, rank = "genus")
```

::: notes
The workflow operates on a TreeSummarizedExperiment. Replace the
example data with your own.
:::

# Alpha diversity {background-color="#13a89e"}

## Alpha diversity: within-sample

::: columns
::: {.column width="45%"}
**Tasks**

1.  Estimate alpha diversity per sample
2.  Draw a histogram
3.  Compare two or more indices, visually and/or statistically
:::

::: {.column width="55%"}
```{r}
#| label: alpha
tse <- addAlpha(
    tse, index = c("shannon", "observed"))

hist(tse$shannon)

plotColData(
    tse, x = "observed", y = "shannon")

cor.test(tse$observed, tse$shannon,
         method = "spearman")
```
:::
:::

::: {.callout-tip}
Richness and evenness capture different things. Do the indices agree?
:::

# Beta diversity {background-color="#f26b5b"}

## Beta diversity: PCoA

Visualize community variation with **PCoA**, and check how the
data transformation affects it:

::: cards
::: {.card .coral}
[Bray–Curtis]{.card-title}

on compositional (relative abundance) data
:::

::: {.card .navy}
[Euclidean]{.card-title}

on CLR-transformed data (Aitchison distance)
:::
:::

```{r}
#| label: pcoa
tse <- runMDS(tse, FUN = getDissimilarity, method = "bray",
              assay.type = "relabundance", name = "PCoA_BC")
tse <- runMDS(tse, FUN = getDissimilarity, method = "euclidean",
              assay.type = "clr", name = "PCoA_CLR")

plotReducedDim(tse, "PCoA_BC",  colour_by = "Diet")
plotReducedDim(tse, "PCoA_CLR", colour_by = "Diet")
```

## Community-level comparisons

Does community composition differ between groups?

::: columns
::: {.column width="45%"}
-   **PERMANOVA** tests group differences in composition
-   Add **covariates** (e.g. age, sex) and see how the results change
-   Check homogeneity of dispersion
:::

::: {.column width="55%"}
```{r}
#| label: permanova
res <- getPERMANOVA(
    tse,
    assay.type = "relabundance",
    method = "bray",
    formula = x ~ Diet + Age,
    test.homogeneity = TRUE)
res
```
:::
:::

::: notes
PERMANOVA is sensitive to differences in group spread (dispersion), not
only location, so check homogeneity of dispersion alongside the test.
:::

# Taxa-level analysis {background-color="#0b2545"}

## Prevalence

What is the most **prevalent** genus in the data?

```{r}
#| label: prevalence
prev <- getPrevalence(
    tse_genus, assay.type = "relabundance",
    detection = 0, sort = TRUE)
head(prev)
```

<br>

[prevalence]{.pill} = fraction of samples where a taxon is detected
above a given abundance threshold

## Core microbiota

Taxa that exceed **0.1 %** relative abundance in **over 50 %** of
samples [@Salonen2012].

::: columns
::: {.column width="55%"}
```{r}
#| label: core
core <- getPrevalent(
    tse_genus,
    assay.type = "relabundance",
    detection = 0.1 / 100,
    prevalence = 50 / 100)
length(core)

tse_core <- tse_genus[core, ]
plotAbundanceDensity(
    tse_core, layout = "jitter",
    assay.type = "relabundance")
```
:::

::: {.column width="45%"}
-   How many core taxa are there?
-   How sensitive is the result to the thresholds?
:::
:::

## Differential abundance

Which genera are associated with a group difference (e.g. sex)?

```{r}
#| label: deseq2
library(DESeq2)

dds <- DESeqDataSet(tse_genus, design = ~ Gender)
dds <- DESeq(dds)
res <- results(dds)
head(res[order(res$padj), ])
```

-   [DESeq2](https://bioconductor.org/packages/DESeq2) [@Love2014];
    for alternatives and benchmarks see the
    [OMA DAA chapter](https://microbiome.github.io/OMA/docs/devel/pages/differential_abundance.html)
-   Compare several methods: results can differ substantially

## Role of covariates

::: statement
Key covariates (diet, medication, age, …) can change the interpretation.
:::

-   Stool consistency, medication and diet explain a considerable part of
    population-level gut microbiome variation [@Falony2016]
-   Include relevant covariates in **PERMANOVA** and **differential
    abundance** models
-   Compare results with and without covariates

# Wrap-up {background-color="#f2b134"}

## Checklist for your report

::: cards
::: {.card .teal}
[Alpha]{.card-title}

Indices, histogram, comparison
:::

::: {.card .coral}
[Beta]{.card-title}

PCoA (BC vs CLR), PERMANOVA with covariates
:::

::: {.card .navy}
[Taxa]{.card-title}

Prevalence, core microbiota, differential abundance
:::

::: {.card .gold}
[Report]{.card-title}

Clear, compact, reproducible
:::
:::

## Resources

-   [Orchestrating Microbiome Analysis (OMA)](https://microbiome.github.io/OMA/)
-   [mia](https://bioconductor.org/packages/mia) and
    [miaViz](https://bioconductor.org/packages/miaViz) documentation
-   [DESeq2 vignette](https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html)

## References
